---
title: "Network Analysis with cograph: Import, Utilities, Centralities & Communities"
description: |
    From raw data to complete network analysis in R --- no boilerplate required
date: "2026-03-05"
author:
  - name: Mohammed Saqr
    url: https://saqr.me
    affiliation: University of Eastern Finland
  - name: Sonsoles López-Pernas
    url: https://sonsoles.me
    affiliation: University of Eastern Finland
citation:
  type: article
  author:
    - family: "Saqr"
      given: "Mohammed"
    - family: "López-Pernas"
      given: "Sonsoles"
categories:
  - tutorial
  - R
  - network analysis
warning: false
message: false
fig.align: center
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    code-tools: false
    theme: cosmo
    self-contained: true
execute:
  warning: false
  message: false
knitr:
  opts_chunk:
    fig.width: 8
    fig.height: 6
    dpi: 150
    comment: ""
---

```{r}
#| label: setup
#| include: false
devtools::load_all(".")
```

# Introduction

Most network analysis in R involves a frustrating ritual: convert your data into the right format, learn a package-specific API, compute one metric at a time, then wrangle outputs back into something usable. If your data comes from a different source --- a transition matrix here, an igraph object there, a statnet network elsewhere --- you spend more time on format conversion than on actual analysis.

**cograph** eliminates this friction. Every function accepts every major network format directly: adjacency matrices, edge-list data frames, igraph, statnet network, qgraph, and tna objects. You never convert manually. You call `centrality_degree(mat)` or `communities(g)` and it works, regardless of what `mat` or `g` happens to be.

This tutorial covers the analysis side of cograph:

1. **Importing networks** --- six formats, zero conversion code
2. **Inspecting and converting** --- getters, summaries, format conversion
3. **Filtering and selecting** --- dplyr-style expressions, centrality-aware selection
4. **Centrality measures** --- 25 node-level measures, edge centrality, directional variants
5. **Network properties** --- 37 summary metrics plus specialized diagnostics
6. **Community detection** --- 11 algorithms, consensus, comparison

For **visualization and plotting**, see the companion tutorial: [Plotting with cograph](cograph-tutorial-plotting.html). For **TNA-specific workflows**, see the [TNA tutorial](../tna-tutorial/index.html).

::: {.callout-note}
## GitHub Development Version

This tutorial covers the full feature set available in the **GitHub development version** of cograph. The CRAN release provides the core plotting engine; the development version adds the full analysis toolkit (centralities, communities, network properties, filtering, and selection).

```{r}
#| label: install
#| eval: false
# Development version with full features
devtools::install_github("sonsoleslp/cograph")
```
:::

# Sample Data

Throughout this tutorial we use a 5-state directed transition matrix representing regulatory learning behaviors. This is the kind of weighted directed network that appears in learning analytics, process mining, and behavioral research.

```{r}
#| label: sample-data
# A 5-state transition matrix (directed, weighted)
states <- c("Explore", "Plan", "Monitor", "Adapt", "Reflect")
mat <- matrix(
  c(0.00, 0.35, 0.20, 0.15, 0.30,
    0.25, 0.00, 0.30, 0.20, 0.25,
    0.15, 0.25, 0.00, 0.35, 0.25,
    0.10, 0.20, 0.35, 0.00, 0.35,
    0.20, 0.15, 0.25, 0.40, 0.00),
  nrow = 5, byrow = TRUE,
  dimnames = list(states, states)
)
round(mat, 2)
```

We also create a larger 8-node undirected network for demonstrations that benefit from more structure:

```{r}
#| label: sample-data-large
# 8-node network for community detection examples
set.seed(42)
nodes8 <- c("A", "B", "C", "D", "E", "F", "G", "H")
mat8 <- matrix(0, 8, 8, dimnames = list(nodes8, nodes8))

# Create two clusters with strong internal connections
# Cluster 1: A, B, C, D
mat8["A", "B"] <- mat8["B", "A"] <- 0.8
mat8["A", "C"] <- mat8["C", "A"] <- 0.6
mat8["B", "C"] <- mat8["C", "B"] <- 0.7
mat8["B", "D"] <- mat8["D", "B"] <- 0.5
mat8["C", "D"] <- mat8["D", "C"] <- 0.6

# Cluster 2: E, F, G, H
mat8["E", "F"] <- mat8["F", "E"] <- 0.9
mat8["E", "G"] <- mat8["G", "E"] <- 0.7
mat8["F", "G"] <- mat8["G", "F"] <- 0.8
mat8["G", "H"] <- mat8["H", "G"] <- 0.6
mat8["F", "H"] <- mat8["H", "F"] <- 0.5

# Weak between-cluster bridges
mat8["D", "E"] <- mat8["E", "D"] <- 0.15
mat8["C", "G"] <- mat8["G", "C"] <- 0.10
```


# Importing Networks

Every cograph function accepts six network formats directly. You never need a separate conversion step --- just pass your data and cograph figures out the rest.

## From a Matrix

The most common input: a square adjacency or weight matrix. Row and column names become node labels. Directedness is auto-detected from matrix symmetry (symmetric = undirected, asymmetric = directed).

```{r}
#| label: import-matrix
# Asymmetric matrix → auto-detected as directed
net <- as_cograph(mat)
net
```

```{r}
#| label: import-matrix-undirected
# Symmetric matrix → auto-detected as undirected
net_u <- as_cograph(mat8)
net_u
```

You can override auto-detection with `directed = TRUE` or `directed = FALSE`.

## From a Data Frame (Edge List)

An edge-list data frame with source, target, and optionally weight columns. Column names are auto-detected from common conventions (`from`/`to`, `source`/`target`, `v1`/`v2`, etc.).

```{r}
#| label: import-edgelist
# Create an edge list data frame
edges_df <- data.frame(
  from   = c("A", "A", "B", "B", "C"),
  to     = c("B", "C", "C", "D", "D"),
  weight = c(0.8, 0.6, 0.7, 0.5, 0.6)
)

net_df <- as_cograph(edges_df)
net_df
```

## From an igraph Object

If you already have an igraph object from another analysis, pass it directly. All vertex and edge attributes are preserved.

```{r}
#| label: import-igraph
library(igraph)

# Resolve namespace conflicts (igraph masks some cograph functions)
communities    <- cograph::communities
membership     <- cograph::membership.cograph_communities
modularity     <- cograph::modularity.cograph_communities
edge_betweenness <- cograph::edge_betweenness
degree_distribution <- cograph::degree_distribution

# Create an igraph object
g <- graph_from_adjacency_matrix(mat, mode = "directed", weighted = TRUE)

# Pass directly --- no conversion needed
net_ig <- as_cograph(g)
net_ig
```

::: {.callout-tip collapse="true"}
## From a statnet network Object

If you work with the `network` / `sna` ecosystem, cograph accepts statnet network objects directly:

```{r}
#| label: import-statnet
#| eval: false
library(network)
n <- network(mat8, directed = FALSE)
net_sn <- as_cograph(n)
```
:::

::: {.callout-tip collapse="true"}
## From a qgraph Object

For users of the `qgraph` package (common in psychological network analysis):

```{r}
#| label: import-qgraph
#| eval: false
library(qgraph)
q <- qgraph(mat)
net_qg <- as_cograph(q)
```
:::

::: {.callout-tip collapse="true"}
## From a TNA Object

cograph has first-class integration with the `tna` package. TNA objects carry additional metadata (initial state probabilities, state labels, color schemes) that cograph preserves automatically:

```{r}
#| label: import-tna
#| eval: false
library(tna)
model <- tna(engagement)
net_tna <- as_cograph(model)
```

When a TNA object is detected, cograph automatically adjusts weight inversion for path-based centrality measures (because TNA transition probabilities are similarities, not distances).
:::

::: {.callout-tip collapse="true"}
## Format Auto-Detection Summary

| Input | Detection Method | Directed? |
|---|---|---|
| Square matrix | `is.matrix()` | Auto from symmetry |
| Data frame | `is.data.frame()` | Auto from bidirectional edges |
| igraph | `inherits(x, "igraph")` | From igraph's `is_directed()` |
| statnet network | `inherits(x, "network")` | From `network::is.directed()` |
| qgraph | `inherits(x, "qgraph")` | From qgraph attributes |
| tna | `inherits(x, "tna")` | Auto from matrix symmetry |

You can always override with `directed = TRUE` or `directed = FALSE`.
:::

# Inspecting Networks

Once you have a cograph_network object, a set of getter functions gives you clean access to every component.

```{r}
#| label: inspect
# Basic properties
cat("Nodes:", n_nodes(net), "\n")
cat("Edges:", n_edges(net), "\n")
cat("Directed:", cograph::is_directed(net), "\n")
cat("Labels:", paste(get_labels(net), collapse = ", "), "\n")
```

```{r}
#| label: inspect-nodes
# Node table
get_nodes(net)
```

```{r}
#| label: inspect-edges
# Edge table (from, to, weight)
get_edges(net)
```

The `summary()` method provides a one-line overview of the network structure:

```{r}
#| label: summary
summary(net)
```

::: {.callout-tip collapse="true"}
## All Getter Functions

| Function | Returns |
|---|---|
| `get_nodes(x)` | Node data frame (id, label, name) |
| `get_edges(x)` | Edge data frame (from, to, weight) |
| `get_labels(x)` | Character vector of node labels |
| `get_source(x)` | Input format name (e.g., "matrix") |
| `get_data(x)` | Original raw input data |
| `get_meta(x)` | Metadata list |
| `n_nodes(x)` | Integer node count |
| `n_edges(x)` | Integer edge count |
| `is_directed(x)` | Logical |
| `get_groups(x)` | Group assignments (if set) |
:::

# Conversion

Sometimes you need to move between formats --- for example, to use an igraph-specific algorithm or to export an edge list to a file. cograph provides four conversion functions that accept **any** input format:

```{r}
#| label: convert-igraph
# Matrix → igraph (for access to igraph's full toolkit)
g <- to_igraph(mat)
g
```

```{r}
#| label: convert-matrix
# igraph → matrix (for matrix algebra or heatmaps)
m <- to_matrix(g)
round(m, 2)
```

```{r}
#| label: convert-df
# Any format → edge list data frame (for CSV export or dplyr pipelines)
df <- to_data_frame(mat)
df
```

::: {.callout-tip collapse="true"}
## Conversion Functions

| Function | Returns | Use Case |
|---|---|---|
| `to_igraph(x)` | igraph object | Access igraph's full algorithm suite |
| `to_matrix(x)` | Adjacency matrix | Matrix algebra, heatmap input |
| `to_data_frame(x)` | Edge list data frame | CSV export, dplyr pipelines |
| `to_network(x)` | statnet network object | Use with sna/ergm packages |

All accept matrix, data.frame, igraph, network, cograph_network, or tna objects. The `directed` parameter can override auto-detection.
:::


# Filtering and Selection

Filtering lets you extract subsets of your network using expressions evaluated against node or edge properties --- including centrality measures computed lazily on demand. This is one of the most powerful features in cograph: you can filter by structural properties without computing them first.

## Filtering Edges

`filter_edges()` uses dplyr-style expressions evaluated against the edge data frame. Keep only the strong transitions:

```{r}
#| label: filter-edges
# Keep only edges with weight > 0.25
strong_net <- filter_edges(mat, weight > 0.25)
get_edges(strong_net)
```

Multiple conditions combine with AND logic:

```{r}
#| label: filter-edges-multi
# Edges from "Adapt" with weight > 0.30
adapt_strong <- filter_edges(mat, from == "Adapt", weight > 0.30)
get_edges(adapt_strong)
```

## Filtering Nodes

`filter_nodes()` evaluates expressions against node properties. What makes this powerful is that **centrality measures are available as lazy-computed columns** --- you don't need to compute them first:

```{r}
#| label: filter-nodes-centrality
# Keep only nodes with degree centrality > 4
# "degree" is computed on the fly inside the filter
high_degree <- filter_nodes(mat, degree > 4)
get_labels(high_degree)
```

::: {.callout-tip collapse="true"}
## Available Lazy Centrality Columns in filter_nodes()

These column names are computed automatically when referenced in a filter expression:

| Column | Measure |
|---|---|
| `degree` | Degree centrality |
| `strength` | Weighted degree |
| `betweenness` | Betweenness centrality |
| `closeness` | Closeness centrality |
| `eigenvector` | Eigenvector centrality |
| `pagerank` | PageRank |
| `hub` | Hub score |
| `authority` | Authority score |
| `coreness` | K-core decomposition |
| `constraint` | Burt's structural constraint |
| `eccentricity` | Eccentricity |
| `transitivity` | Local clustering coefficient |

These are not pre-computed --- they are evaluated only when you reference them, so there is no performance cost when filtering by other attributes.
:::

## Selecting Nodes

`select_nodes()` provides six selection modes that combine with AND logic:

```{r}
#| label: select-by-name
# Select specific nodes by name
sub <- select_nodes(mat, name = c("Plan", "Monitor", "Adapt"))
get_labels(sub)
```

```{r}
#| label: select-top
# Top 3 nodes by betweenness centrality
top3 <- select_nodes(mat, top = 3, by = "betweenness")
get_labels(top3)
```

```{r}
#| label: select-neighbors
# Select "Monitor" and its direct neighbors
neighborhood <- select_neighbors(mat, of = "Monitor", order = 1)
get_labels(neighborhood)
```

::: {.callout-tip collapse="true"}
## All Selection Modes

| Mode | Parameter | Example |
|---|---|---|
| By name | `name = c("A", "B")` | Exact node names |
| By index | `index = c(1, 3, 5)` | Positional indices |
| Top-N by centrality | `top = 5, by = "degree"` | Highest centrality |
| Neighbors | `neighbors_of = "A", order = 2` | Ego network to depth 2 |
| Component | `component = "largest"` | Largest connected component |
| NSE expression | `... (e.g., degree > 5)` | Any expression, like `filter_nodes()` |

Combine modes freely --- they are intersected with AND logic.
:::

## Selecting Edges

```{r}
#| label: select-top-edges
# Top 5 strongest edges
top5 <- select_top_edges(mat, n = 5)
get_edges(top5)
```

```{r}
#| label: select-bridges
# Bridge edges (whose removal disconnects components)
bridges <- select_bridges(mat8)
get_edges(bridges)
```

::: {.callout-tip collapse="true"}
## Edge Selection Modes

| Mode | Parameter | Description |
|---|---|---|
| Top-N | `top = 10, by = "weight"` | Strongest/highest edges |
| Involving nodes | `involving = c("A", "B")` | Edges touching specific nodes |
| Between sets | `between = list(set1, set2)` | Edges between two node groups |
| Bridges only | `bridges_only = TRUE` | Edges whose removal disconnects |
| Mutual only | `mutual_only = TRUE` | Reciprocated edges (directed networks) |

**Lazy edge metrics** are available in expressions: `abs_weight`, `from_degree`, `to_degree`, `edge_betweenness`, `is_bridge`, `is_mutual`, `same_community`.
:::

## Convenience Wrappers

For the most common operations, cograph provides one-call wrappers:

| Function | Equivalent |
|---|---|
| `select_top(x, n, by)` | `select_nodes(x, top = n, by = by)` |
| `select_neighbors(x, of, order)` | `select_nodes(x, neighbors_of = of, order = order)` |
| `select_component(x, "largest")` | `select_nodes(x, component = "largest")` |
| `select_bridges(x)` | `select_edges(x, bridges_only = TRUE)` |
| `select_top_edges(x, n)` | `select_edges(x, top = n)` |
| `select_edges_involving(x, nodes)` | `select_edges(x, involving = nodes)` |
| `select_edges_between(x, s1, s2)` | `select_edges(x, between = list(s1, s2))` |


# Centrality Measures

Centrality reduces each node to a single number reflecting its structural importance. Different measures answer different questions, and the right measure depends on what "important" means for your research question.

## The centrality() Function

`centrality()` computes multiple measures at once and returns a tidy data frame:

```{r}
#| label: centrality-all
# Compute the most common measures at once
cent <- centrality(mat, measures = c(
  "degree", "strength", "betweenness", "closeness",
  "eigenvector", "pagerank", "harmonic", "hub", "authority"
))
cent
```

Request specific measures:

```{r}
#| label: centrality-specific
# Just degree and betweenness
cent_subset <- centrality(mat, measures = c("degree", "betweenness", "pagerank"))
cent_subset
```

::: {.callout-tip collapse="true"}
## Key Parameters

| Parameter | Default | What It Controls |
|---|---|---|
| `measures` | `"all"` | Which measures to compute (character vector or "all") |
| `mode` | `"all"` | Direction for mode-aware measures: `"all"`, `"in"`, `"out"` |
| `normalized` | `FALSE` | Normalize to [0, 1]? |
| `weighted` | `TRUE` | Use edge weights? |
| `directed` | `NULL` | Override directedness (NULL = auto-detect) |
| `damping` | `0.85` | PageRank damping factor |
| `alpha` | `1` | Alpha centrality exponent |
| `k` | `3` | k-reach neighborhood depth |
| `digits` | `NULL` | Round output (NULL = no rounding) |
| `sort_by` | `NULL` | Sort rows by a measure name |
:::

## Individual Wrappers

Every measure has a dedicated function that returns a named numeric vector. These are simpler when you need just one measure:

```{r}
#| label: centrality-wrappers
# Each returns a named numeric vector
centrality_degree(mat)
centrality_betweenness(mat)
centrality_pagerank(mat)
```

## Directional Measures

For directed networks, ten measures support `mode = "in"` and `mode = "out"`. cograph provides shorthand wrappers so you don't need the `mode` argument:

```{r}
#| label: centrality-directional
# InStrength: total incoming weight (which states are "attractors"?)
centrality_instrength(mat)

# OutStrength: total outgoing weight (which states are frequent origins?)
centrality_outstrength(mat)
```

::: {.callout-tip collapse="true"}
## All 25 Centrality Measures

| Measure | Directional? | Question It Answers |
|---|---|---|
| **degree** | in/out/all | How many connections does this node have? |
| **strength** | in/out/all | How much total weight flows through this node? |
| **betweenness** | --- | Does this node sit on paths between other nodes? |
| **closeness** | in/out/all | How quickly can this node reach (or be reached by) others? |
| **eigenvector** | --- | Is this node connected to other well-connected nodes? |
| **pagerank** | --- | How important is this node in a random-walk model? |
| **authority** | --- | Is this node a destination for important hubs? |
| **hub** | --- | Does this node point to important authorities? |
| **eccentricity** | in/out/all | How far is the farthest reachable node? |
| **coreness** | in/out/all | What is the densest core this node belongs to? |
| **constraint** | --- | How constrained is this node by its local neighborhood? |
| **transitivity** | --- | Do this node's neighbors connect to each other? |
| **harmonic** | in/out/all | Closeness variant robust to disconnected networks |
| **diffusion** | --- | How well does this node spread activation? |
| **leverage** | --- | Does this node have more connections than its neighbors? |
| **kreach** | --- | How many nodes are reachable within k steps? |
| **alpha** | --- | Bonacich alpha (exogenous influence) centrality |
| **power** | --- | Bonacich power centrality |
| **subgraph** | --- | Importance based on all closed walks through this node |
| **laplacian** | --- | Importance in the graph's Laplacian spectrum |
| **load** | --- | Load centrality (traffic-based betweenness variant) |
| **current_flow_closeness** | --- | Information centrality via electrical current analogy |
| **current_flow_betweenness** | --- | Betweenness via current flow (not just shortest paths) |
| **voterank** | --- | Iterative influence maximization score |
| **percolation** | --- | Importance for network connectivity under failure |

**Directional wrappers**: `centrality_indegree()`, `centrality_outdegree()`, `centrality_instrength()`, `centrality_outstrength()`, `centrality_incloseness()`, `centrality_outcloseness()`, `centrality_inharmonic()`, `centrality_outharmonic()`, `centrality_ineccentricity()`, `centrality_outeccentricity()`.
:::

## Edge Centrality

Node centrality tells you which nodes are important. Edge centrality tells you which **transitions** are important --- which connections serve as structural bridges.

```{r}
#| label: edge-centrality
# Edge betweenness: which edges carry the most shortest-path traffic?
eb <- edge_betweenness(mat)
eb
```

```{r}
#| label: edge-centrality-df
# Full edge centrality data frame
edf <- edge_centrality(mat)
edf
```

## TNA Weight Inversion

When the input is a TNA object, cograph automatically inverts weights for path-based measures (betweenness, closeness, harmonic, eccentricity, and others). This is because TNA transition probabilities are **similarities** (higher = closer) while shortest-path algorithms expect **distances** (lower = closer). You don't need to do anything --- it just works.


# Network Properties

While centrality characterizes individual nodes, network properties characterize the **network as a whole**: how dense it is, how clustered, how efficiently information flows through it.

## network_summary()

The `network_summary()` function computes up to 37 metrics in a single call:

```{r}
#| label: summary-basic
# Basic summary (18 metrics)
network_summary(mat)
```

```{r}
#| label: summary-extended
# Extended summary (adds girth, radius, efficiency, etc.)
network_summary(mat, extended = TRUE)
```

```{r}
#| label: summary-detailed
# Detailed summary (adds centrality distribution statistics)
network_summary(mat, detailed = TRUE)
```

::: {.callout-tip collapse="true"}
## What Each Metric Tells You

**Always included (18 metrics):**

| Metric | Interpretation |
|---|---|
| `node_count`, `edge_count` | Network size |
| `density` | Proportion of possible edges that exist |
| `component_count` | Number of disconnected components |
| `diameter` | Longest shortest path in the network |
| `mean_distance` | Average shortest path length |
| `min_cut` | Minimum edges to disconnect the network |
| `centralization_degree` | How unequal is the degree distribution? |
| `centralization_closeness` | How centralized is reachability? |
| `centralization_betweenness` | How much does one node dominate bridging? |
| `centralization_eigenvector` | How concentrated is influence? |
| `transitivity` | Global clustering coefficient |
| `reciprocity` | Proportion of mutual edges (directed) |
| `assortativity_degree` | Do high-degree nodes connect to other high-degree nodes? |
| `hub_score`, `authority_score` | Max hub/authority centrality |
| `is_connected`, `is_bipartite` | Structural properties |

**Extended (`extended = TRUE`, adds 8):**

| Metric | Interpretation |
|---|---|
| `girth` | Shortest cycle length |
| `radius` | Minimum eccentricity |
| `vertex_connectivity` | Min nodes to remove to disconnect |
| `largest_clique_size` | Size of the largest complete subgraph |
| `cut_vertex_count` | Articulation points (bridges) |
| `bridge_count` | Bridge edges |
| `global_efficiency` | Average inverse shortest path |
| `local_efficiency` | Mean ego-network efficiency |

**Detailed (`detailed = TRUE`, adds 11):**

Descriptive statistics (mean, SD, median) for degree, strength, betweenness, closeness, eigenvector, PageRank, constraint, and local transitivity.
:::

## Individual Property Functions

For specific analyses, dedicated functions provide richer output than the summary table:

```{r}
#| label: degree-dist
# Degree distribution (histogram of degree values)
degree_distribution(mat)
```

```{r}
#| label: small-world
# Small-world coefficient (compares clustering and path length to random graphs)
sw <- network_small_world(mat8, n_random = 100)
cat("Small-world coefficient:", round(sw, 3), "\n")
```

```{r}
#| label: rich-club
# Rich-club coefficient (do high-degree nodes preferentially connect?)
rc <- network_rich_club(mat8, normalized = TRUE, n_random = 100)
cat("Rich-club coefficient:", round(rc, 3), "\n")
```

::: {.callout-tip collapse="true"}
## All Individual Property Functions

| Function | Returns | Use Case |
|---|---|---|
| `degree_distribution(x)` | Numeric vector | Degree histogram, power-law testing |
| `network_girth(x)` | Integer | Shortest cycle length |
| `network_radius(x)` | Integer | Network compactness |
| `network_vertex_connectivity(x)` | Integer | Robustness to node failure |
| `network_clique_size(x)` | Integer | Largest complete subgraph |
| `network_cut_vertices(x)` | Character vector | Articulation points |
| `network_bridges(x)` | Data frame | Bridge edges |
| `network_global_efficiency(x)` | Numeric [0, 1] | Information flow efficiency |
| `network_local_efficiency(x)` | Numeric [0, 1] | Local fault tolerance |
| `network_small_world(x)` | List (sigma, omega) | Small-world characterization |
| `network_rich_club(x, k)` | Data frame | Rich-club phenomenon |
:::


# Community Detection

Communities are groups of nodes that are more densely connected to each other than to the rest of the network. Detecting them reveals the **modular structure** of a system: which behaviors cluster together, which brain regions form functional modules, which actors form cohesive subgroups.

## Basic Usage

The `communities()` function applies a community detection algorithm and returns a `cograph_communities` object:

```{r}
#| label: communities-basic
# Detect communities using the default algorithm (Louvain)
comms <- communities(mat8)
comms
```

```{r}
#| label: communities-membership
# Which community does each node belong to?
membership(comms)
```

```{r}
#| label: communities-sizes
# How large is each community?
community_sizes(comms)
```

```{r}
#| label: communities-modularity
# Modularity score (how well-separated are the communities?)
modularity(comms)
```

## Choosing an Algorithm

cograph provides 11 community detection algorithms. Each has different strengths:

```{r}
#| label: communities-algorithms
# Fast greedy: modularity optimization (fast, undirected)
comms_fg <- community_fast_greedy(mat8)

# Walktrap: random-walk based (captures flow structure)
comms_wt <- community_walktrap(mat8)

# Leiden: improved Louvain with resolution parameter
comms_ld <- community_leiden(mat8)
```

::: {.callout-tip collapse="true"}
## All 11 Algorithms

| Function | Short Alias | Algorithm | Best For |
|---|---|---|---|
| `community_louvain()` | `com_lv()` | Louvain modularity | General purpose, large networks |
| `community_leiden()` | `com_ld()` | Leiden (improved Louvain) | Better-defined communities |
| `community_fast_greedy()` | `com_fg()` | Fast greedy | Quick modularity optimization |
| `community_walktrap()` | `com_wt()` | Random walk | Flow-based structure |
| `community_infomap()` | `com_im()` | Information flow | Directed flow networks |
| `community_label_propagation()` | `com_lp()` | Label propagation | Very large networks |
| `community_edge_betweenness()` | `com_eb()` | Edge betweenness | Small networks, clear bridges |
| `community_leading_eigenvector()` | `com_le()` | Leading eigenvector | Spectral decomposition |
| `community_spinglass()` | `com_sg()` | Spin glass | Small, fine-grained structure |
| `community_optimal()` | `com_op()` | Exact optimization | Tiny networks (NP-hard) |
| `community_fluid()` | `com_fl()` | Fluid communities | When k is known a priori |

Every function also has a **two-letter alias** (e.g., `com_lv()`, `com_fg()`) for quick interactive use.
:::

## Consensus Communities

A single algorithm run can be sensitive to initialization. Consensus communities run the algorithm many times and keep only the assignments that appear consistently:

```{r}
#| label: communities-consensus
# Run Louvain 100 times, keep consistent assignments
consensus <- community_consensus(mat8, method = "louvain", n_runs = 100, seed = 42)
consensus
```

## Comparing Community Structures

When you run two different algorithms, how similar are their results? `compare_communities()` quantifies this with information-theoretic metrics:

```{r}
#| label: communities-compare
# Compare Louvain vs. Walktrap
nmi <- compare_communities(comms, comms_wt, method = "nmi")
cat("Normalized Mutual Information:", round(nmi, 3), "\n")

# Adjusted Rand Index
ari <- compare_communities(comms, comms_wt, method = "adjusted.rand")
cat("Adjusted Rand Index:", round(ari, 3), "\n")
```

::: {.callout-tip collapse="true"}
## Comparison Methods

| Method | Range | Interpretation |
|---|---|---|
| `"nmi"` | [0, 1] | 1 = identical, 0 = unrelated |
| `"vi"` | [0, log(n)] | Lower = more similar (Variation of Information) |
| `"rand"` | [0, 1] | Proportion of concordant pairs |
| `"adjusted.rand"` | [-1, 1] | Chance-corrected Rand (0 = random) |
| `"split.join"` | [0, 2n] | Lower = more similar |
:::

## Visualizing Communities

The `plot()` method for community objects colors nodes by community membership:

```{r}
#| label: fig-communities-plot
#| fig-cap: "Community structure detected by Louvain"
#| fig-width: 8
#| fig-height: 8
plot(comms, net_u)
```

You can also use `color_communities()` to get a color vector for use with `splot()`:

```{r}
#| label: color-communities
# Get community colors as a named vector (for splot's node_fill parameter)
colors <- color_communities(mat8, method = "louvain")
colors
```


# Putting It All Together

A complete analysis pipeline, from import to communities, in a few lines:

```{r}
#| label: workflow
#| eval: false
devtools::load_all(".")

# 1. Import (works with any format)
net <- as_cograph(my_matrix)

# 2. Inspect
summary(net)
n_nodes(net)
n_edges(net)

# 3. Filter
core <- filter_nodes(net, degree > 3)
strong <- filter_edges(net, weight > 0.2)

# 4. Select subnetwork
top5 <- select_top(net, n = 5, by = "betweenness")
ego <- select_neighbors(net, of = "Monitor", order = 2)

# 5. Centrality
centrality(net, measures = c("degree", "betweenness", "pagerank"))
centrality_instrength(net)
centrality_outstrength(net)

# 6. Network properties
network_summary(net, extended = TRUE, detailed = TRUE)
network_small_world(net)

# 7. Communities
comms <- communities(net)
membership(comms)
community_sizes(comms)
plot(comms, net)

# 8. Export
to_igraph(net)                # for igraph analyses
to_data_frame(net)            # for CSV export
write.csv(centrality(net), "centralities.csv")
```

# References {.unnumbered}

- cograph: Modern R Package for Network Visualization. <https://github.com/sonsoleslp/cograph>

- Saqr, M., López-Pernas, S., Törmänen, T., Kaliisa, R., Misiejuk, K., & Tikka, S. (2025). Transition Network Analysis: A Novel Framework for Modeling, Visualizing, and Identifying the Temporal Patterns of Learners and Learning Processes. In *Proceedings of the 15th International Learning Analytics and Knowledge Conference (LAK '25)* (pp. 351--361). ACM. <https://doi.org/10.1145/3706468.3706513>

- Tikka, S., López-Pernas, S., & Saqr, M. (2025). tna: An R Package for Transition Network Analysis. *Applied Psychological Measurement*. <https://doi.org/10.1177/01466216251348840>
