---
title: "Motif Analysis with cograph: Patterns, Significance & Visualization"
description: |
    Discover recurring interaction patterns in networks --- from aggregate census to individual-level significance testing
date: "2026-03-13"
author:
  - name: Mohammed Saqr
    url: https://saqr.me
    affiliation: University of Eastern Finland
  - name: Sonsoles López-Pernas
    url: https://sonsoles.me
    affiliation: University of Eastern Finland
citation:
  type: article
  author:
    - family: "Saqr"
      given: "Mohammed"
    - family: "López-Pernas"
      given: "Sonsoles"
categories:
  - tutorial
  - R
  - network analysis
  - motifs
warning: false
message: false
fig.align: center
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    code-tools: false
    theme: cosmo
    self-contained: true
execute:
  warning: false
  message: false
knitr:
  opts_chunk:
    fig.width: 8
    fig.height: 6
    dpi: 150
    comment: ""
---

```{r}
#| label: setup
#| include: false
library(igraph)
library(tna)
devtools::load_all(".")
```

# Introduction

Networks encode *pairwise* relationships --- who connects to whom and how strongly. But many of the most interesting behavioral patterns are *higher-order*: three nodes forming a feedback loop, a hub funneling information to two recipients, or a fully reciprocal clique. These three-node subgraph patterns are called **triadic motifs**, and they reveal the micro-level building blocks of network structure.

**Why motifs matter:**

- **Feed-forward loops** (A $\to$ B $\to$ C with A $\to$ C) indicate hierarchical information flow
- **Reciprocal cycles** (A $\leftrightarrow$ B $\leftrightarrow$ C $\leftrightarrow$ A) suggest tightly coupled feedback
- **Stars** (A $\to$ B, A $\to$ C) reveal broadcasting or gatekeeping roles
- **Cliques** (all six directed edges present) mark the densest possible local structure

cograph provides two main functions for motif analysis:

| Function | Mode | What it answers |
|----------|------|-----------------|
| `motifs()` | **Census** | How many of each MAN type? (nodes exchangeable) |
| `subgraphs()` | **Instances** | Which specific node triples form each pattern? (nodes identified) |

Both functions:

- Work with **any input**: tna objects, matrices, igraph, cograph networks, or edge list data frames
- **Auto-detect** session/actor columns for individual-level analysis
- Support **windowing** (rolling or tumbling) for temporal analysis
- Run **significance testing** by default using the exact configuration model

This tutorial covers:

1. **The MAN classification** --- 16 triad types and what they mean
2. **Census mode** --- `motifs()` for type-level counts with significance
3. **Instance mode** --- `subgraphs()` for specific node triples
4. **Individual-level analysis** --- per-session motif frequencies
5. **Pattern filtering** --- triangles, networks, closed patterns
6. **Visualization** --- bar plots, significance plots, network diagrams, abstract patterns
7. **Auto-detection and windowing** --- edge list metadata and temporal windows
8. **Comparing groups** --- motif profiles across conditions

For **network visualization**, see the companion tutorial: [Plotting with cograph](cograph-tutorial-plotting.html). For **centralities and communities**, see [Network Analysis with cograph](cograph-tutorial-analysis.html).

::: {.callout-note}
## GitHub Development Version

This tutorial covers features available in the **GitHub development version** of cograph.

```{r}
#| label: install
#| eval: false
# Development version with full features
devtools::install_github("sonsoleslp/cograph")
```
:::

# The Data

We use the `coding` dataset from the [tna](https://cran.r-project.org/package=tna) package throughout this tutorial. It contains 429 coding sessions where pairs of students collaborate on programming tasks. Each session is a sequence of interaction behaviors:

```{r}
#| label: data-overview
library(tna)

# Build a TNA model from the coding dataset
Mod <- tna(coding)

cat("States:", paste(Mod$labels, collapse = ", "), "\n")
cat("Number of states:", length(Mod$labels), "\n")
cat("Number of sessions:", nrow(coding), "\n")
```

The nine behavioral states capture different interaction modes during collaborative coding --- from giving commands to making requests, verifying work, or expressing frustration.

```{r}
#| label: fig-network
#| fig-cap: "Transition network of collaborative coding behaviors. Edge widths reflect transition probabilities."
#| fig-width: 7
#| fig-height: 7
splot(Mod, minimum = 0.04, title = "Coding Interactions")
```

The weight matrix shows transition probabilities between states:

```{r}
#| label: weight-matrix
round(Mod$weights, 3)
```

# The MAN Classification

Every set of three nodes in a directed network forms one of exactly **16 possible triad configurations**, classified by the MAN (Mutual-Asymmetric-Null) system (Holland & Leinhardt, 1976):

- **M** (Mutual): number of reciprocal (bidirectional) pairs
- **A** (Asymmetric): number of one-way pairs
- **N** (Null): number of unconnected pairs

The three digits are followed by a letter suffix for subtypes: **C** (cycle), **T** (transitive), **D** (down/out), **U** (up/in).

::: {.callout-tip collapse="true"}
## Full MAN Type Reference

| Type | Name | Description |
|------|------|-------------|
| 003 | Empty | No edges between any pair |
| 012 | Single edge | One directed edge only |
| 102 | Mutual pair | One reciprocal pair, third node isolated |
| 021D | Out-star | A $\to$ B, A $\to$ C (one source, two targets) |
| 021U | In-star | B $\to$ A, C $\to$ A (two sources, one target) |
| 021C | Chain | A $\to$ B $\to$ C (sequential path) |
| 111D | Out-star + mutual | Mutual pair plus outgoing edge |
| 111U | In-star + mutual | Mutual pair plus incoming edge |
| 030T | Feed-forward | A $\to$ B $\to$ C, A $\to$ C (transitive) |
| 030C | Cycle | A $\to$ B $\to$ C $\to$ A (directed cycle) |
| 201 | Double mutual | Two mutual pairs sharing a node |
| 120D | Regulated out-star | Mutual pair plus two outgoing edges |
| 120U | Regulated in-star | Mutual pair plus two incoming edges |
| 120C | Regulated cycle | Mutual pair plus cycle through third |
| 210 | Near-clique | All pairs connected, one asymmetric |
| 300 | Clique | All three pairs fully mutual |
:::

The first four types (003, 012, 102, 021D/U/C) describe sparse or sequential patterns. The last seven (030T/C through 300) involve triangles --- all three node pairs have at least one edge. Triangle motifs are typically the most informative because they capture genuine three-way interaction structure.

# Census Mode: `motifs()`

The `motifs()` function counts MAN type frequencies --- how many triads of each type exist. Significance testing is **on by default**, comparing observed counts against a configuration model null.

## From a Matrix (Aggregate)

```{r}
#| label: census-matrix
mat <- Mod$weights

# Aggregate census with significance (default)
m <- motifs(mat, n_perm = 500, seed = 42)
m
```

::: {.callout-tip}
## Interpreting a Dense Network

When all states are interconnected (as in this transition matrix where every state can transition to every other), you often see all triads as type 300 (clique). Motif analysis becomes more informative when you either:

1. **Threshold the matrix** to keep only strong transitions
2. **Use the "expected" edge method** to focus on disproportionately strong edges
3. **Work at the individual level** where sessions have sparser patterns
:::

## From a tna Object (Individual-Level)

When given a tna object, `motifs()` automatically performs **individual-level** analysis --- counting type frequencies across all 429 sessions:

```{r}
#| label: census-tna
m_ind <- motifs(Mod, n_perm = 20, seed = 42)
m_ind
```

The `level: individual` and `n_units` fields confirm per-session analysis. Significance uses the exact configuration model: stub-shuffling preserves exact in-degree and out-degree sequences per session.

## Without Significance

For a quick count without permutation testing:

```{r}
#| label: census-nosig
m_fast <- motifs(Mod, significance = FALSE)
m_fast
```

## Plotting the Census

```{r}
#| label: fig-census-types
#| fig-cap: "Distribution of MAN triad types across all sessions."
#| fig-width: 8
#| fig-height: 5
plot(m_fast, type = "types")
```

```{r}
#| label: fig-census-sig
#| fig-cap: "Significance plot showing over- and under-represented triad types."
#| fig-width: 8
#| fig-height: 5
plot(m_ind, type = "significance")
```

```{r}
#| label: fig-census-patterns
#| fig-cap: "Abstract MAN triad patterns showing the structure of each type."
#| fig-width: 10
#| fig-height: 8
plot(m_fast, type = "patterns")
```

# Instance Mode: `subgraphs()`

While `motifs()` tells you *how many* of each type exist, `subgraphs()` tells you *which specific nodes* form each pattern. It is a convenience wrapper for `motifs(..., named_nodes = TRUE)`.

```{r}
#| label: subgraphs-basic
sg <- subgraphs(Mod, significance = FALSE)
sg
```

Each row shows a specific triad (e.g., "Command - Request - Specify") with its MAN type and how many sessions contain it. The `observed` column counts the number of individuals (sessions) exhibiting that triad.

## With Significance Testing

Significance is on by default. It tests whether each specific triad appears in more sessions than expected under the configuration model:

```{r}
#| label: sig-testing
#| cache: true
sg_sig <- subgraphs(Mod, n_perm = 100, seed = 42)
sg_sig
```

Significant triads (`sig = TRUE`) appear in more or fewer sessions than the null model predicts.

## Plotting Subgraphs

### Network Diagrams (Default)

```{r}
#| label: fig-triads-network
#| fig-cap: "Top triads as network diagrams. Each panel shows one triad with its MAN type, session count, z-score, and p-value."
#| fig-width: 10
#| fig-height: 8
plot(sg_sig, n = 15)
```

### Customizing Appearance

```{r}
#| label: fig-triads-custom
#| fig-cap: "Customized triad network diagrams with adjusted sizes and colors."
#| fig-width: 10
#| fig-height: 6
plot(sg_sig, n = 10, ncol = 5,
     node_size = 6, label_size = 8, title_size = 9,
     color = "#2166AC")
```

### Significance Plot

```{r}
#| label: fig-significance
#| fig-cap: "Significance plot showing over-represented (red) and under-represented (blue) triads."
#| fig-width: 8
#| fig-height: 6
plot(sg_sig, type = "significance", n = 15)
```

::: {.callout-tip collapse="true"}
## How the Permutation Test Works

For each permutation:

1. Within each session, the transition matrix is randomized using the **exact configuration model** --- row stubs and column stubs are created from the degree sequence, then column stubs are randomly shuffled. This preserves exact in-degree and out-degree per session.
2. Triads are counted in each randomized session
3. The total count across sessions is recorded

After all permutations, the null distribution gives the expected count and standard deviation. The z-score measures how far the observed count is from the null mean. A two-tailed p-value tests whether the triad is significantly over- *or* under-represented.

The `n_perm` parameter controls precision. Use 100 for exploration, 1000 for publication-quality results.
:::

# Pattern Filtering

Both `motifs()` and `subgraphs()` support four pattern filters:

```{r}
#| label: pattern-filters
# Triangle: all 3 pairs connected (default)
m_tri <- motifs(Mod, significance = FALSE, pattern = "triangle")
cat("Triangle types:", nrow(m_tri$results), "types,", sum(m_tri$results$count), "total\n")

# Network: exclude empty and simple chains
m_net <- motifs(Mod, significance = FALSE, pattern = "network")
cat("Network types:", nrow(m_net$results), "types,", sum(m_net$results$count), "total\n")

# All: include everything
m_all <- motifs(Mod, significance = FALSE, pattern = "all")
cat("All types:", nrow(m_all$results), "types,", sum(m_all$results$count), "total\n")
```

::: {.callout-tip collapse="true"}
## Which Pattern Filter Should I Use?

- **`"triangle"`** (default): Best for most analyses. Focuses on genuine three-way interaction where all three node pairs have at least one edge.
- **`"network"`**: Broader view that also includes stars (021D, 021U) and mutual pairs.
- **`"closed"`**: Like network but excludes certain open patterns.
- **`"all"`**: Complete census of all 16 types. Good for exploratory analysis.
:::

## Include/Exclude Specific Types

```{r}
#| label: include-exclude
# Only feed-forward loops and cycles
m_ff <- motifs(Mod, include = c("030T", "030C"), significance = FALSE)
cat("Feed-forward + cycle:", sum(m_ff$results$count), "total\n")

# Triangles but exclude cliques
m_noclique <- motifs(Mod, exclude = "300", significance = FALSE)
cat("No cliques:", sum(m_noclique$results$count), "total\n")
```

# Auto-Detection and Windowing

When working with edge list data, `motifs()` and `subgraphs()` can **automatically detect** session/actor grouping columns and ordering columns.

## Auto-Detection

```{r}
#| label: auto-detect
# Create synthetic edge list with standard column names
set.seed(42)
states <- c("Plan", "Execute", "Monitor", "Adapt")
edge_list <- do.call(rbind, lapply(1:5, function(s) {
  data.frame(
    from = sample(states, 15, replace = TRUE),
    to = sample(states, 15, replace = TRUE),
    session_id = paste0("s", s),
    order = seq_len(15),
    stringsAsFactors = FALSE
  )
}))

net <- as_cograph(edge_list)

# motifs() auto-detects session_id column → individual-level
m_auto <- motifs(net, significance = FALSE, min_transitions = 1)
cat("Level:", m_auto$level, "\n")
cat("Units (sessions):", m_auto$n_units, "\n")
```

Standard column names detected automatically: `session_id`, `session`, `actor`, `user`, `participant`, `individual`, `id` (in priority order, case-insensitive).

## Override with `actor`

```{r}
#| label: actor-override
# Add a project column
edge_list$project <- ifelse(edge_list$session_id %in% c("s1", "s2", "s3"), "p1", "p2")
net2 <- as_cograph(edge_list)

# Override auto-detection to group by project instead
m_proj <- motifs(net2, actor = "project", significance = FALSE, min_transitions = 1)
cat("Units (projects):", m_proj$n_units, "\n")
```

## Windowing

Split each actor's transitions into temporal windows:

```{r}
#| label: windowing
# Tumbling windows of size 5 (non-overlapping)
m_tumble <- motifs(net, window = 5, window_type = "tumbling",
                   significance = FALSE, min_transitions = 1)
cat("Tumbling window units:", m_tumble$n_units, "\n")

# Rolling windows of size 5 (overlapping)
m_roll <- motifs(net, window = 5, window_type = "rolling",
                 significance = FALSE, min_transitions = 1)
cat("Rolling window units:", m_roll$n_units, "\n")
```

# Working with Different Input Types

`motifs()` and `subgraphs()` accept all standard input types:

## From a Matrix

```{r}
#| label: matrix-input
states5 <- c("Plan", "Execute", "Monitor", "Adapt", "Reflect")
mat5 <- matrix(c(
  0.00, 0.35, 0.20, 0.15, 0.30,
  0.25, 0.00, 0.30, 0.20, 0.25,
  0.15, 0.25, 0.00, 0.35, 0.25,
  0.10, 0.20, 0.35, 0.00, 0.35,
  0.20, 0.15, 0.25, 0.40, 0.00
), nrow = 5, byrow = TRUE,
dimnames = list(states5, states5))

# Census
m5 <- motifs(mat5, n_perm = 200, seed = 42)
m5

# Instances with node labels
sg5 <- subgraphs(mat5, significance = FALSE, pattern = "all", min_count = NULL)
head(sg5$results[, c("triad", "type", "observed")])
```

## From an igraph Object

```{r}
#| label: igraph-input
g <- igraph::graph_from_adjacency_matrix(mat5, mode = "directed", weighted = TRUE)

# Same functions work with igraph
m_ig <- motifs(g, n_perm = 200, seed = 42)

# Results match the matrix input
cat("Z-scores match:", all.equal(m5$results$z, m_ig$results$z), "\n")
```

## From a Data Frame (Edge List)

```{r}
#| label: df-input
# Direct data.frame input (no as_cograph needed)
el <- data.frame(
  from = c("A", "B", "A", "C", "B", "C", "A", "B"),
  to   = c("B", "C", "C", "A", "A", "B", "C", "A"),
  session = c("s1", "s1", "s1", "s1", "s2", "s2", "s2", "s2"),
  stringsAsFactors = FALSE
)

m_df <- motifs(el, significance = FALSE, min_transitions = 1)
cat("Level:", m_df$level, "| Units:", m_df$n_units, "\n")
```

# Comparing Groups

A powerful application is comparing motif profiles across groups.

## Side-by-Side Analysis

```{r}
#| label: group-comparison
#| cache: true
# Split coding dataset into two groups
sessions <- rownames(coding)
projects <- sub("_S[0-9]+$", "", sessions)
project_ids <- unique(projects)

set.seed(42)
group_assign <- ifelse(
  projects %in% project_ids[1:floor(length(project_ids) / 2)],
  "Group_A", "Group_B"
)

coding_a <- coding[group_assign == "Group_A", ]
coding_b <- coding[group_assign == "Group_B", ]

Mod_a <- tna(coding_a)
Mod_b <- tna(coding_b)

cat("Group A:", nrow(coding_a), "sessions\n")
cat("Group B:", nrow(coding_b), "sessions\n")
```

```{r}
#| label: group-motifs
# Census per group
m_a <- motifs(Mod_a, significance = TRUE, n_perm = 100, seed = 42)
m_b <- motifs(Mod_b, significance = TRUE, n_perm = 100, seed = 42)

cat("=== Group A ===\n")
print(m_a$results)

cat("\n=== Group B ===\n")
print(m_b$results)
```

## Type Distribution Comparison

```{r}
#| label: fig-group-types
#| fig-cap: "Motif type distributions for Group A (left) and Group B (right)."
#| fig-width: 12
#| fig-height: 5
par(mfrow = c(1, 2))
barplot(m_a$type_summary, main = "Group A: Type Distribution",
        col = "#4FC3F7", las = 2, cex.names = 0.8)
barplot(m_b$type_summary, main = "Group B: Type Distribution",
        col = "#fbb550", las = 2, cex.names = 0.8)
par(mfrow = c(1, 1))
```

## Instance-Level Group Comparison

```{r}
#| label: group-subgraphs
# Top subgraphs per group
sg_a <- subgraphs(Mod_a, significance = FALSE, top = 10)
sg_b <- subgraphs(Mod_b, significance = FALSE, top = 10)

cat("=== Group A: Top triads ===\n")
print(sg_a$results[, c("triad", "type", "observed")])

cat("\n=== Group B: Top triads ===\n")
print(sg_b$results[, c("triad", "type", "observed")])
```

# Edge Method Options

The `edge_method` parameter controls how edges are determined from weighted transition matrices:

```{r}
#| label: edge-methods
# "any" (default): edge exists if count > 0
m_any <- motifs(Mod, edge_method = "any", pattern = "all", significance = FALSE)
cat("Edge method 'any':", sum(m_any$results$count), "total triads\n")

# "expected": edge exists if observed/expected >= threshold
m_exp <- motifs(Mod, edge_method = "expected", edge_threshold = 1.5,
                pattern = "all", significance = FALSE)
cat("Edge method 'expected' (1.5x):", sum(m_exp$results$count), "total triads\n")
```

::: {.callout-tip collapse="true"}
## Choosing an Edge Method

- **`"any"`**: Most permissive. An edge exists if there is *any* transition. Good for sparse individual-level data.
- **`"expected"`**: An edge exists only if it occurs `threshold` times more than expected by chance. Good for dense networks.
- **`"percent"`**: An edge exists if it accounts for at least `threshold` proportion of all transitions from that node.

For individual-level analysis with tna objects, `"any"` is typically best. For aggregate matrices, `"expected"` helps cut through density.
:::

# Legacy Functions

The following functions are still available for backward compatibility but `motifs()` and `subgraphs()` are the recommended API:

| Legacy Function | Replacement |
|----------------|-------------|
| `triad_census(mat)` | `motifs(mat, significance = FALSE, pattern = "all")` |
| `motif_census(mat)` | `motifs(mat)` (significance on by default) |
| `extract_triads(mat)` | `subgraphs(mat, pattern = "all", min_count = NULL)` |
| `extract_motifs(Mod)` | `subgraphs(Mod)` |

# Summary

| Function | Mode | Default | Best For |
|----------|------|---------|----------|
| `motifs(x)` | Census | `significance = TRUE`, `pattern = "triangle"` | How many of each type? Over/under-represented? |
| `subgraphs(x)` | Instances | `significance = TRUE`, `min_count = 5` | Which specific node triples? |

: {tbl-colwidths="[15,12,38,35]"}

**Recommended workflow:**

1. **Census** with `motifs(Mod)` to identify which types are over/under-represented
2. **Instances** with `subgraphs(Mod)` to see which specific nodes form those patterns
3. **Visualize** with `plot(result, type = ...)` for triads, types, significance, or patterns
4. **Compare groups** by running `motifs()` on each group's tna model

::: {.callout-note}
## References

- Holland, P.W. & Leinhardt, S. (1976). Local structure in social networks. *Sociological Methodology*, 7, 1--45.
- Milo, R., Shen-Orr, S., Itzkovitz, S., et al. (2002). Network motifs: Simple building blocks of complex networks. *Science*, 298(5594), 824--827.
- Wasserman, S. & Faust, K. (1994). *Social Network Analysis: Methods and Applications*. Cambridge University Press.
:::
