---
title: "Network Visualization with cograph: A Complete Plotting Guide"
description: |
    Publication-ready network plots with one function call --- from basic graphs to donut-pie composites
date: "2026-03-05"
author:
  - name: Mohammed Saqr
    url: https://saqr.me
    affiliation: University of Eastern Finland
  - name: Sonsoles López-Pernas
    url: https://sonsoles.me
    affiliation: University of Eastern Finland
citation:
  type: article
  author:
    - family: "Saqr"
      given: "Mohammed"
    - family: "López-Pernas"
      given: "Sonsoles"
categories:
  - tutorial
  - R
  - visualization
warning: false
message: false
fig.align: center
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    code-tools: false
    theme: cosmo
    self-contained: true
execute:
  warning: false
  message: false
knitr:
  opts_chunk:
    fig.width: 8
    fig.height: 8
    dpi: 150
    comment: ""
---

```{r}
#| label: setup
#| include: false
library(tna)
devtools::load_all(".")
```

# Introduction

Network visualization in R has traditionally meant choosing between two extremes: quick-and-dirty plots that look like hairballs, or heavily customized code that takes longer to write than the analysis itself. **cograph** collapses this tradeoff. The `splot()` function produces publication-ready network visualizations from a single call, with sensible defaults that you can override parameter by parameter.

This tutorial demonstrates everything through a single TNA model of collaborative regulation behaviors. We start with a zero-argument plot and progressively layer in customization --- so you can see exactly what each parameter changes.

1. **The plot** --- one line, zero arguments, publication-ready
2. **Node styling** --- shapes, colors, sizes, labels, centrality-based scaling
3. **Edge styling** --- width, color, curvature, labels, thresholds
4. **Donuts and pies** --- composite node shapes for multivariate data
5. **Themes and palettes** --- pre-built visual themes
6. **Pipe-friendly API** --- the `sn_*` function family for chainable plot building
7. **Specialized plots** --- chord diagrams, heatmaps, alluvial flows
8. **Group comparisons** --- difference networks and multi-group grids
9. **Bootstrap** --- validated network visualization
10. **Permutation testing** --- significant group differences
11. **Mixed networks** --- directed and undirected edges combined

For **network analysis** (centrality, communities, filtering), see the companion tutorial: [Network Analysis with cograph](cograph-tutorial-analysis.html). For TNA-specific workflows, see the [TNA tutorial](../tna-tutorial/index.html).

::: {.callout-note}
## GitHub Development Version

This tutorial covers the full feature set available in the **GitHub development version** of cograph. The CRAN release provides the core plotting engine; the development version adds themes, pipe API, specialized plots, and additional features.

```{r}
#| label: install
#| eval: false
# Development version with full features
devtools::install_github("sonsoleslp/cograph")
```
:::


# Building the Model

We use the built-in `group_regulation` dataset from the `tna` package --- coded collaborative regulation behaviors from student groups. This gives us a 9-state directed transition network, the kind of rich structure that showcases cograph's capabilities.

```{r}
#| label: build-model
library(tna)
data("group_regulation")
model <- tna(group_regulation)
model
```

We also prepare a group model for later comparison and permutation sections:

```{r}
#| label: build-group-model
data("group_regulation_long")
prepared <- prepare_data(group_regulation_long,
                         action = "Action", actor = "Actor", time = "Time")
group_model <- group_tna(prepared, group = "Achiever")
```


# The Plot

One function. Zero arguments beyond the model. Publication-ready.

```{r}
#| label: fig-the-plot
#| fig-cap: "splot(model) --- that's it"
splot(model)
```

When cograph receives a TNA object, it auto-applies TNA-appropriate defaults: oval layout, dark blue edges weighted by transition probability, donut rings showing initial state probabilities, and directional arrows. Every aspect of this plot can be customized --- the rest of this tutorial shows how.


# Node Styling

## Shapes

Every node can have its own shape. cograph supports circles, squares, hexagons, diamonds, triangles, pentagons, stars, hearts, and more:

```{r}
#| label: fig-node-shapes
#| fig-cap: "Different shapes per node"
splot(model,
  node_shape = c("circle", "square", "hexagon", "diamond", "triangle",
                 "pentagon", "star", "heart", "circle"),
  layout = "circle"
)
```

::: {.callout-tip collapse="true"}
## All Built-In Shapes

`"circle"`, `"square"`, `"triangle"`, `"diamond"`, `"pentagon"`, `"hexagon"`, `"star"`, `"heart"`, `"ellipse"`, `"cross"`, `"rectangle"`, `"pie"`, `"donut"`

You can also register **custom SVG shapes** with `register_svg_shape()` and use them by name.
:::

## Colors and Sizes

```{r}
#| label: fig-node-colors
#| fig-cap: "Custom fills, borders, and per-node sizes"
splot(model,
  node_fill = palette_viridis(9),
  node_border_color = "white",
  node_border_width = 2,
  node_size = c(12, 8, 10, 9, 7, 11, 8, 10, 9)
)
```

## Scaling Nodes by Centrality

`scale_nodes_by` maps a centrality measure to node size --- larger nodes are more central. This is one of the most useful parameters for making network structure immediately visible:

```{r}
#| label: fig-scale-nodes
#| fig-cap: "Node size proportional to different centrality measures"
#| fig-width: 10
#| fig-height: 5
par(mfrow = c(1, 2), mar = c(1, 1, 2, 1))
splot(model, scale_nodes_by = "betweenness", node_size_range = c(3, 14),
      title = "Betweenness")
splot(model, scale_nodes_by = "strength", node_size_range = c(3, 14),
      title = "Strength")
```

::: {.callout-tip collapse="true"}
## scale_nodes_by Options

Any centrality measure name works: `"degree"`, `"strength"`, `"betweenness"`, `"closeness"`, `"eigenvector"`, `"pagerank"`, `"authority"`, `"hub"`, `"eccentricity"`, `"coreness"`, `"constraint"`, `"transitivity"`, `"harmonic"`, `"diffusion"`, `"leverage"`, `"kreach"`.

For directional measures (in/out), use a list: `scale_nodes_by = list("strength", mode = "in")`.

| Parameter | Default | Description |
|---|---|---|
| `scale_nodes_by` | `NULL` | Centrality measure name or list with mode |
| `node_size_range` | `c(2, 8)` | `c(min, max)` size range for mapping |
:::


# Edge Styling

## Width and Curvature

```{r}
#| label: fig-edge-width
#| fig-cap: "Custom edge width range with curvature"
splot(model,
  edge_width = 3,
  curvature = 0.3,
  arrow_size = 0.015
)
```

## Edge Labels

Display transition probabilities directly on edges:

```{r}
#| label: fig-edge-labels
#| fig-cap: "Edge labels showing transition weights"
splot(model,
  edge_labels = TRUE,
  edge_label_size = 0.6,
  edge_label_bg = "white",
  threshold = 0.05
)
```

## Thresholding

Hide weak transitions to reveal the backbone:

```{r}
#| label: fig-threshold
#| fig-cap: "Thresholding removes visual clutter"
#| fig-width: 10
#| fig-height: 5
par(mfrow = c(1, 2), mar = c(1, 1, 2, 1))
splot(model, title = "All edges")
splot(model, threshold = 0.10, title = "threshold = 0.10")
```

::: {.callout-tip collapse="true"}
## Edge Parameters

| Parameter | Default | Description |
|---|---|---|
| `edge_width` | auto | Base edge width |
| `edge_width_range` | auto | `c(min, max)` width range |
| `edge_color` | auto | Edge color |
| `edge_alpha` | `1` | Transparency |
| `edge_style` | `"solid"` | `"solid"`, `"dashed"`, `"dotted"` |
| `curvature` | auto | Curve amount |
| `arrow_size` | auto | Arrowhead size |
| `threshold` | `0` | Hide edges below this weight |
| `edge_labels` | `FALSE` | Show weight labels |
| `edge_label_style` | `"plain"` | `"plain"`, `"ci"`, `"stars"`, `"full"` |
| `edge_positive_color` | auto | Color for positive weights |
| `edge_negative_color` | auto | Color for negative weights |
:::


# Donuts and Pies

Donut and pie node shapes encode **additional quantitative information** directly in the node rendering. In TNA, the donut ring is automatically filled from initial state probabilities --- but you can use it for any node-level metric.

## Donut Nodes

The default TNA plot already shows donut fills from `model$inits`. You can customize the appearance:

```{r}
#| label: fig-donuts-custom
#| fig-cap: "Hexagonal donut shapes with custom color"
splot(model,
  donut_shape = "hexagon",
  donut_color = "#2E7D32"
)
```

## Pie Nodes

Pie chart nodes show a categorical distribution per node. You can combine them with the outer donut ring:

```{r}
#| label: fig-donut-pie
#| fig-cap: "Donut ring (initial probability) + inner pie segments"
# Each node gets a 3-slice pie (e.g., from a clustering or condition split)
pie_vals <- list(
  c(0.5, 0.3, 0.2), c(0.4, 0.4, 0.2), c(0.3, 0.3, 0.4),
  c(0.6, 0.2, 0.2), c(0.2, 0.5, 0.3), c(0.4, 0.3, 0.3),
  c(0.7, 0.2, 0.1), c(0.3, 0.4, 0.3), c(0.5, 0.2, 0.3)
)

splot(model,
  pie_values = pie_vals,
  pie_colors = c("#1976D2", "#FFA000", "#C62828"),
  layout = "circle"
)
```

::: {.callout-tip collapse="true"}
## Donut & Pie Parameters

| Parameter | Default | Description |
|---|---|---|
| `donut_fill` | auto (from TNA inits) | Numeric 0--1 per node |
| `donut_color` | auto | Fill color for the ring |
| `donut_inner_ratio` | `0.5` | Inner radius ratio (0--1) |
| `donut_shape` | `"circle"` | `"circle"`, `"square"`, `"hexagon"`, `"diamond"`, `"triangle"`, `"pentagon"` |
| `donut_show_value` | `FALSE` | Show numeric value in center |
| `donut_value_suffix` | `""` | Text after value (e.g., `"%"`) |
| `pie_values` | `NULL` | List of numeric vectors per node |
| `pie_colors` | auto | Colors for pie segments |
:::


# Themes and Palettes

Themes apply a coordinated set of visual defaults with a single parameter:

```{r}
#| label: fig-themes
#| fig-cap: "Four built-in themes"
#| fig-width: 10
#| fig-height: 10
par(mfrow = c(2, 2), mar = c(1, 1, 2, 1))
splot(model, theme = "classic",    title = "classic")
splot(model, theme = "dark",       title = "dark")
splot(model, theme = "minimal",    title = "minimal")
splot(model, theme = "colorblind", title = "colorblind")
```

```{r}
#| label: fig-palettes
#| fig-cap: "Color palettes for nodes"
#| fig-width: 10
#| fig-height: 5
par(mfrow = c(1, 3), mar = c(1, 1, 2, 1))
splot(model, node_fill = palette_viridis(9),    title = "viridis")
splot(model, node_fill = palette_pastel(9),     title = "pastel")
splot(model, node_fill = palette_colorblind(9), title = "colorblind")
```

::: {.callout-tip collapse="true"}
## All Themes

| Theme | Description |
|---|---|
| `"classic"` | Clean, professional (default) |
| `"dark"` | Dark background, light elements |
| `"minimal"` | Reduced chrome, subtle colors |
| `"colorblind"` | Colorblind-safe palette |
| `"grayscale"` | No color, grayscale only |
| `"vibrant"` | Bold, saturated colors |
| `"nature"` | Earth-tone palette |
| `"viridis"` | Viridis color scale |

Use `list_themes()` to see all registered themes. Create custom themes with `register_theme()`.
:::


# Pipe-Friendly API

The `sn_*` function family lets you build plots incrementally using the pipe operator. Each function takes a cograph_network and returns a modified one:

```{r}
#| label: fig-pipe
#| fig-cap: "Dark-themed hexagon network built with pipes"
model |>
  cograph(layout = "circle") |>
  sn_theme("dark") |>
  sn_nodes(size = 10, shape = "hexagon",
           border_color = "white", border_width = 2) |>
  sn_edges(width = 4, curvature = 0.2) |>
  soplot(title = "Built with pipes")
```

::: {.callout-tip collapse="true"}
## All sn_* Functions

| Function | What It Configures |
|---|---|
| `sn_layout(network, layout, seed)` | Layout algorithm |
| `sn_theme(network, theme)` | Visual theme |
| `sn_palette(network, palette, target, by)` | Color palette for nodes or edges |
| `sn_nodes(network, ...)` | All node aesthetics (size, shape, fill, labels, pie, donut) |
| `sn_edges(network, ...)` | All edge aesthetics (width, color, labels, curvature, CI) |
| `sn_render(network, ...)` | Render with grid graphics (alias for `soplot()`) |
| `sn_ggplot(network, title)` | Convert to ggplot2 object |
| `sn_save(network, filename, ...)` | Save to file (PDF/PNG/SVG) |
:::


# Specialized Plots

## Chord Diagram

Chord diagrams show flows as curved ribbons around a circle. The width of each ribbon is proportional to the transition weight:

```{r}
#| label: fig-chord
#| fig-cap: "Chord diagram of transition flows"
plot_chord(model, chord_alpha = 0.6, title = "Transition Flows")
```

::: {.callout-tip collapse="true"}
## Chord Diagram Parameters

| Parameter | Default | Description |
|---|---|---|
| `segment_colors` | auto | Colors for each node's arc |
| `chord_color_by` | `"source"` | Color ribbons by `"source"`, `"target"`, or `"weight"` |
| `chord_alpha` | `0.5` | Ribbon transparency |
| `self_loop` | `FALSE` | Show self-transition arcs? |
| `threshold` | `0` | Hide chords below this weight |
:::

## Heatmap

The adjacency matrix as a color-coded grid, optionally grouped by cluster:

```{r}
#| label: fig-heatmap
#| fig-cap: "Transition heatmap with clustered groups"
#| fig-width: 9
#| fig-height: 8
clusters <- list(
  Social    = c("discuss", "consensus", "cohesion", "emotion"),
  Cognitive = c("plan", "monitor", "adapt", "synthesis"),
  Meta      = c("coregulate")
)

plot_heatmap(model,
  cluster_list = clusters,
  show_values = TRUE,
  value_size = 3.5,
  title = "Transition Heatmap"
)
```

## Alluvial / Sankey Diagram

Shows source-to-target flow as an alluvial diagram:

```{r}
#| label: fig-alluvial
#| fig-cap: "Alluvial diagram of transition flows"
#| fig-width: 9
#| fig-height: 7
plot_alluvial(model$weights, threshold = 0.05)
```


# Group Comparisons

## Multi-Group Grid

When you pass a `group_tna` object, splot renders all groups in a grid:

```{r}
#| label: fig-group-grid
#| fig-cap: "High vs Low achiever networks side by side"
#| fig-width: 12
#| fig-height: 6
splot(group_model, i = NULL)
```

## Difference Network

`plot_compare()` computes and visualizes the element-wise difference between two groups. Green edges are stronger in the first group; red edges are stronger in the second:

```{r}
#| label: fig-compare
#| fig-cap: "Difference network: High - Low achievers"
plot_compare(group_model,
  pos_color = "#2A9D8F",
  neg_color = "#E76F51",
  title = "High vs Low"
)
```


# Bootstrap

The bootstrap procedure determines which edges are stable properties of the data and which are sampling artifacts. Pass a `tna_bootstrap` object to `splot()` and it automatically styles significant vs. non-significant edges:

```{r}
#| label: bootstrap
#| cache: true
set.seed(265)
boot <- bootstrap(model, iter = 1000, level = 0.05)
```

```{r}
#| label: fig-bootstrap-styled
#| fig-cap: "Bootstrap-validated network (solid = significant, dashed = non-significant)"
splot(boot)
```

```{r}
#| label: fig-bootstrap-sig
#| fig-cap: "Only significant edges shown"
splot(boot, display = "significant")
```

```{r}
#| label: fig-bootstrap-ci
#| fig-cap: "Confidence interval labels on edges"
splot(boot, display = "ci", threshold = 0.05)
```

::: {.callout-tip collapse="true"}
## Bootstrap Display Modes

| `display` | Description |
|---|---|
| `"styled"` (default) | Significant = solid, non-significant = dashed |
| `"significant"` | Only significant edges shown |
| `"full"` | All edges shown with equal styling |
| `"ci"` | Confidence interval labels on edges |

Additional parameters: `show_stars` (significance stars), `show_ci` (CI values), `width_by` (scale width by weight or CI width).
:::


# Permutation Testing

Permutation testing evaluates whether between-group differences are statistically significant. Pass the result to `splot()` to visualize which edge differences survived the permutation test:

```{r}
#| label: permutation
#| cache: true
set.seed(265)
perm <- permutation_test(group_model, iter = 1000)
```

```{r}
#| label: fig-permutation
#| fig-cap: "Permutation test: significant between-group edge differences"
splot(perm)
```

::: {.callout-tip collapse="true"}
## Permutation Plot Parameters

| Parameter | Default | Description |
|---|---|---|
| `show_nonsig` | `TRUE` | Show non-significant edges? |
| `edge_nonsig_color` | `"gray80"` | Color for non-significant edges |
| `edge_nonsig_alpha` | `0.3` | Transparency for non-significant edges |
| `show_stars` | `FALSE` | Show significance stars |
| `show_effect` | `FALSE` | Show effect sizes |
:::


# Mixed Networks

Overlay directed and undirected edges on the same plot. This is useful when you have two types of relationships --- for example, symmetric co-occurrence patterns and asymmetric transition probabilities. The defaults follow TNA conventions: oval layout, weight-scaled edges, dotted edge starts, and visible edge labels.

```{r}
#| label: fig-mixed
#| fig-cap: "Co-occurrence (steel blue, straight) and transitions (dark blue, curved) with TNA-style defaults"
# Undirected: symmetric co-occurrence (half the weight for visual balance)
set.seed(42)
sym <- (model$weights + t(model$weights)) / 2
noise <- matrix(runif(length(sym), -0.02, 0.02), nrow = nrow(sym))
noise <- (noise + t(noise)) / 2  # keep symmetric
diag(noise) <- 0
sym <- pmax(sym * 0.5 + noise, 0)

# Directed: original transition probabilities
asym <- model$weights

# Defaults: oval layout, blue edges, weight-scaled widths, edge labels
plot_mixed_network(sym, asym, threshold = 0.03,
  title = "Co-occurrence + Transitions"
)
```

You can distinguish edge types by color: undirected co-occurrence edges are straight steel-blue lines, while directed transitions are curved dark-blue arrows with dotted starts. Edge widths scale automatically by weight, so stronger connections stand out.

```{r}
#| label: fig-mixed-custom
#| fig-cap: "Custom colors to visually separate edge types"
plot_mixed_network(sym, asym, threshold = 0.03,
  sym_color = "#E63946",
  asym_color = "#003355",
  title = "Red undirected + Blue directed"
)
```


# Saving and Exporting

## Direct File Output

```{r}
#| label: save-splot
#| eval: false
# Save as PNG
splot(model, filename = "network.png", width = 8, height = 8, res = 300)

# Save as PDF
splot(model, filename = "network.pdf", width = 8, height = 8)

# Save as SVG
splot(model, filename = "network.svg", width = 8, height = 8)
```

## Pipe-Friendly Saving

```{r}
#| label: save-pipe
#| eval: false
model |>
  cograph(layout = "circle") |>
  sn_theme("dark") |>
  sn_nodes(size = 10) |>
  sn_save("network_dark.png", width = 8, height = 8, dpi = 300)
```

## ggplot2 Export

```{r}
#| label: save-ggplot
#| eval: false
p <- model |>
  cograph(layout = "circle") |>
  sn_ggplot()

ggplot2::ggsave("network.pdf", p, width = 8, height = 8)
```


# Complete Parameter Reference

::: {.callout-tip collapse="true"}
## splot() --- All Parameters

**Input & Layout:**
`x`, `layout`, `directed`, `seed`, `theme`, `title`, `background`, `groups`

**Node Appearance:**
`node_size`, `node_size2`, `node_shape`, `node_fill`, `node_border_color`, `node_border_width`, `node_alpha`, `scale_nodes_by`, `node_size_range`

**Labels:**
`labels`, `label_size`, `label_color`, `label_position`, `label_fontface`, `label_fontfamily`

**Donut:**
`donut_fill`, `donut_color`, `donut_inner_ratio`, `donut_shape`, `donut_show_value`, `donut_value_suffix`, `donut_value_digits`, `donut2_values`, `donut2_colors`

**Pie:**
`pie_values`, `pie_colors`

**Edge Appearance:**
`edge_color`, `edge_width`, `edge_size`, `edge_width_range`, `edge_scale_mode`, `edge_cutoff`, `edge_alpha`, `edge_style`, `edge_start_style`, `edge_positive_color`, `edge_negative_color`, `curvature`, `curves`

**Arrows:**
`arrow_size`, `arrow_angle`, `show_arrows`, `bidirectional`

**Edge Labels:**
`edge_labels`, `edge_label_size`, `edge_label_color`, `edge_label_bg`, `edge_label_position`, `edge_label_offset`, `edge_label_fontface`, `edge_label_style`, `edge_label_template`

**CI Underlay:**
`edge_ci`, `edge_ci_scale`, `edge_ci_alpha`, `edge_ci_color`, `edge_ci_style`

**Weight / Threshold:**
`threshold`, `minimum`, `maximum`, `weight_digits`

**Layout Control:**
`rescale`, `layout_scale`, `layout_margin`, `aspect`, `loop_rotation`

**File Output:**
`filename`, `filetype`, `width`, `height`, `res`

**TNA:**
`i` (group index for group_tna)
:::

::: {.callout-tip collapse="true"}
## All Available Layouts

**Built-in:** `"oval"` (default), `"circle"`, `"spring"`, `"grid"`, `"random"`, `"star"`, `"bipartite"`, `"groups"`

**igraph (two-letter codes):** `"kk"` (Kamada-Kawai), `"fr"` (Fruchterman-Reingold), `"drl"`, `"mds"`, `"lgl"`, `"ni"` (nicely), `"tr"` (tree), `"st"` (Sugiyama)

You can also pass a **custom layout matrix** (2-column matrix with x, y coordinates) or a **layout function**.
:::


# References {.unnumbered}

- cograph: Modern R Package for Network Visualization. <https://github.com/sonsoleslp/cograph>

- Saqr, M., López-Pernas, S., Törmänen, T., Kaliisa, R., Misiejuk, K., & Tikka, S. (2025). Transition Network Analysis: A Novel Framework for Modeling, Visualizing, and Identifying the Temporal Patterns of Learners and Learning Processes. In *Proceedings of the 15th International Learning Analytics and Knowledge Conference (LAK '25)* (pp. 351--361). ACM. <https://doi.org/10.1145/3706468.3706513>

- Tikka, S., López-Pernas, S., & Saqr, M. (2025). tna: An R Package for Transition Network Analysis. *Applied Psychological Measurement*. <https://doi.org/10.1177/01466216251348840>
