---
title: "Network Robustness and Backbone Extraction with cograph"
description: |
    Simulate targeted attacks and random failures to assess network resilience ---
    with principled backbone extraction via the disparity filter
date: "2026-03-13"
author:
  - name: Mohammed Saqr
    url: https://saqr.me
    affiliation: University of Eastern Finland
  - name: Sonsoles López-Pernas
    url: https://sonsoles.me
    affiliation: University of Eastern Finland
citation:
  type: article
  author:
    - family: "Saqr"
      given: "Mohammed"
    - family: "López-Pernas"
      given: "Sonsoles"
categories:
  - tutorial
  - R
  - network analysis
  - robustness
warning: false
message: false
fig.align: center
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    code-tools: false
    theme: cosmo
    self-contained: true
execute:
  warning: false
  message: false
knitr:
  opts_chunk:
    fig.width: 8
    fig.height: 6
    dpi: 150
    comment: ""
---

```{r}
#| label: setup
#| include: false
devtools::load_all(".")
library(tna)
library(igraph)
library(ggplot2)
```

# Introduction

How robust is a network to disruption? If the most central node is removed, does the entire structure collapse --- or does it absorb the shock? **Network robustness analysis** answers this by simulating sequential removal of nodes (or edges) and tracking how the largest connected component shrinks (Albert, Jeong & Barabási, 2000).

This tutorial covers two cograph capabilities:

1. **Backbone extraction** with `disparity_filter()` --- a statistically principled method for identifying significant edges (Serrano, Boguñá & Vespignani, 2009)
2. **Robustness analysis** with `robustness()` --- targeted attacks and random failures on the resulting backbone

We use two real-world datasets: a **student interaction network** (edge list) and a **human-AI behavioral network** (transition matrix from 429 programming sessions).

::: {.callout-note}
## GitHub Development Version

Robustness analysis and the disparity filter are available in the **GitHub development version** of cograph.

```{r}
#| label: install
#| eval: false
devtools::install_github("sonsoleslp/cograph")
```
:::

# Backbone Extraction with the Disparity Filter

## Why Not Simple Thresholds?

A common preprocessing step is to threshold edge weights (e.g., "remove all edges below 0.10"). This applies a global cutoff that ignores network structure --- a weight of 0.05 might be trivial for a well-connected hub but critical for a peripheral node with only two neighbors.

The **disparity filter** (Serrano et al., 2009) is a statistically principled alternative. For each node $i$ with degree $k_i$ and each edge $(i, j)$ with normalized weight $p_{ij} = w_{ij}/s_i$, the null-model p-value is:

$$\alpha_{ij} = (1 - p_{ij})^{(k_i - 1)}$$

An edge is retained in the backbone if $\alpha_{ij} < \text{level}$ from **either** endpoint. This preserves locally important edges even in low-traffic regions of the network.

## Quick Example

```{r}
#| label: disparity-demo
mat_demo <- matrix(c(
  0.0, 0.8, 0.1, 0.1,
  0.3, 0.0, 0.5, 0.2,
  0.1, 0.1, 0.0, 0.8,
  0.2, 0.1, 0.7, 0.0
), nrow = 4, byrow = TRUE)
rownames(mat_demo) <- colnames(mat_demo) <- c("A", "B", "C", "D")

# Significance matrix (1 = backbone edge, 0 = filtered)
disparity_filter(mat_demo, level = 0.50)
```

The function accepts matrices, igraph objects, tna objects, and cograph_network objects. When called on a tna or igraph object, it returns a `tna_disparity` object with the backbone, original weights, and summary statistics.

# Robustness Analysis

## The Algorithm

`robustness()` simulates sequential removal of nodes (or edges) and tracks the size of the largest connected component at each step.

Three **measures** determine *which* element to remove next:

| Measure | Target | Deterministic? |
|---|---|---|
| **Betweenness** | Bridge nodes between network regions | Yes |
| **Degree** | Highly connected hub nodes | Yes |
| **Random** | Random order, averaged over `n_iter` runs | No |

Two **strategies** determine *when* centrality is computed:

| Strategy | Description | Reference |
|---|---|---|
| **Sequential** (default) | Recalculate centrality after each removal --- a stronger, adaptive attack | Holme et al. (2002) |
| **Static** | Compute centrality once on the original network, remove in that fixed order | Albert et al. (2000); brainGraph |

::: {.callout-tip}
## Which Strategy to Choose?

**Sequential** recalculation produces a stronger attack because removing a hub changes which nodes become the new bridges. Use it as the default. **Static** ordering matches the original Albert et al. (2000) methodology and the brainGraph R package --- use it when comparing against those benchmarks.
:::

## Quick Example

```{r}
#| label: robustness-demo
# Scale-free network (robust to random, vulnerable to targeted)
set.seed(42)
g_demo <- igraph::sample_pa(50, m = 2, directed = FALSE)

rob_btw <- robustness(g_demo, measure = "betweenness")
rob_deg <- robustness(g_demo, measure = "degree")
rob_rnd <- robustness(g_demo, measure = "random", n_iter = 100, seed = 42)

plot_robustness(rob_btw, rob_deg, rob_rnd,
                title = "Scale-free Network: Vertex Robustness")
```

The gap between the targeted-attack curves (red/green) and the random-failure curve (blue) is a hallmark of scale-free networks.

# Part 1: Student Interaction Network

## Loading and Building the Network

The `student_interactions` dataset is an edge list of interactions between 34 students during collaborative learning. The same pair can appear multiple times, so the raw graph is a **multigraph**.

```{r}
#| label: student-load
data(student_interactions)
el <- student_interactions[student_interactions$from != student_interactions$to, ]

cat("Interactions:", nrow(el), "\n")
cat("Unique student pairs:", nrow(unique(el)), "\n")
cat("Students:", length(unique(c(el$from, el$to))), "\n")
```

We build three network representations:

```{r}
#| label: student-build
# 1. Multigraph: raw parallel edges
g_multi <- graph_from_data_frame(el, directed = FALSE)

# 2. Aggregated: simple graph (edge count = number of interactions)
g_agg <- simplify(g_multi)

# 3. Weighted: parallel edges collapsed to weights
E(g_multi)$weight <- 1
g_weighted <- simplify(g_multi, edge.attr.comb = list(weight = "sum"))

cat("Multigraph: ", vcount(g_multi), "nodes,", ecount(g_multi), "edges\n")
cat("Aggregated: ", vcount(g_agg), "nodes,", ecount(g_agg), "edges\n")
cat("Weighted:   ", vcount(g_weighted), "nodes,", ecount(g_weighted),
    "edges, max weight =", max(E(g_weighted)$weight), "\n")
```

## Extracting the Backbone

```{r}
#| label: student-backbone
disp_student <- disparity_filter(g_weighted, level = 0.50)
disp_student
```

```{r}
#| label: fig-student-network
#| fig-cap: "Student interaction network: full weighted graph vs disparity backbone"
#| fig-height: 6
#| fig-width: 14
par(mfrow = c(1, 2))

splot(g_weighted, layout = "fr", title = "Full Network")

# Backbone: remove isolated nodes for cleaner visualization
mat_bb <- disp_student$weights_filtered
keep <- rowSums(mat_bb) > 0 | colSums(mat_bb) > 0
splot(mat_bb[keep, keep], layout = "fr",
      title = sprintf("Backbone (p < 0.50, %d/%d nodes)",
                       sum(keep), nrow(mat_bb)))
par(mfrow = c(1, 1))
```

## Robustness: Effect of Backbone Extraction

Does removing statistically insignificant edges change the robustness profile? The backbone strips away weak ties, exposing structural vulnerabilities that the full network masks.

```{r}
#| label: fig-student-backbone-robust
#| fig-cap: "Backbone extraction reveals hidden vulnerability to targeted attacks"
#| fig-height: 5
#| fig-width: 12
ggplot_robustness(
  "Full network" = g_agg,
  "Backbone (p < 0.50)" = disp_student$weights_filtered,
  n_iter = 100, seed = 42, ncol = 2
)
```

In the full network, weak ties keep the graph connected even as hubs are removed. The backbone tells a different story: targeted attacks collapse the network much faster than random failures.

## Robustness: Effect of Aggregation

Does collapsing parallel edges to a simple graph matter?

```{r}
#| label: fig-student-aggregation
#| fig-cap: "Vertex robustness is largely unaffected by aggregation"
#| fig-height: 5
#| fig-width: 12
ggplot_robustness(
  "Multigraph (raw)" = g_multi,
  "Aggregated (simple)" = g_agg,
  n_iter = 100, seed = 42, ncol = 2
)
```

For **vertex** robustness, both produce similar curves --- the largest connected component depends on node reachability, not edge multiplicity.

## Edge Robustness

When we remove **edges** instead of nodes, aggregation matters much more --- a multigraph has far more edges to remove before connectivity breaks.

```{r}
#| label: fig-student-edge
#| fig-cap: "Edge robustness: aggregation matters for edge removal"
#| fig-height: 5
#| fig-width: 12
rob_edge_multi_btw <- robustness(g_multi, type = "edge", measure = "betweenness")
rob_edge_multi_rnd <- robustness(g_multi, type = "edge", measure = "random",
                                  n_iter = 100, seed = 42)
rob_edge_agg_btw <- robustness(g_agg, type = "edge", measure = "betweenness")
rob_edge_agg_rnd <- robustness(g_agg, type = "edge", measure = "random",
                                n_iter = 100, seed = 42)

par(mfrow = c(1, 2))
plot_robustness(rob_edge_multi_btw, rob_edge_multi_rnd,
                title = "Multigraph: Edge Robustness")
plot_robustness(rob_edge_agg_btw, rob_edge_agg_rnd,
                title = "Aggregated: Edge Robustness")
par(mfrow = c(1, 1))
```

## Summary Metrics

`robustness_summary()` returns the **AUC** (area under the robustness curve; 1.0 = maximally robust) and **critical thresholds** --- the fraction of nodes removed before the largest component drops below 50% and 10% of its original size.

```{r}
#| label: tbl-student-summary
#| tbl-cap: "Student Network: Full vs Backbone Robustness"
rob_full_btw <- robustness(g_agg, measure = "betweenness")
rob_full_deg <- robustness(g_agg, measure = "degree")
rob_full_rnd <- robustness(g_agg, measure = "random", n_iter = 100, seed = 42)

rob_bb_btw <- robustness(disp_student$weights_filtered, measure = "betweenness")
rob_bb_deg <- robustness(disp_student$weights_filtered, measure = "degree")
rob_bb_rnd <- robustness(disp_student$weights_filtered, measure = "random",
                          n_iter = 100, seed = 42)

knitr::kable(rbind(
  cbind(Network = "Full", robustness_summary(rob_full_btw, rob_full_deg, rob_full_rnd)),
  cbind(Network = "Backbone", robustness_summary(rob_bb_btw, rob_bb_deg, rob_bb_rnd))
), row.names = FALSE)
```


# Part 2: Human-AI Interaction Network

## The Dense Network Problem

The `human_ai` dataset contains 429 programming sessions coded into 17 behavioral states. We build a transition network using `tna::tna()`.

```{r}
#| label: hai-setup
mod <- tna(human_ai)
mat_full <- mod$weights
n_states <- nrow(mat_full)
n_possible <- n_states * (n_states - 1)

cat("States:  ", n_states, "\n")
cat("Sessions:", nrow(human_ai), "\n")
cat("Edges:   ", sum(mat_full > 0), "of", n_possible, "possible\n")
cat("Density: ", round(sum(mat_full > 0) / n_possible, 2), "\n")
```

At over 90% density, removing any single node cannot fragment the network. Robustness analysis on the raw matrix produces near-identical straight lines for all attack strategies --- there is no structural vulnerability to exploit. We need backbone extraction.

## Backbone Extraction

```{r}
#| label: hai-backbone
disp_hai <- disparity_filter(mod, level = 0.50)
disp_hai

mat_backbone <- disp_hai$weights_filtered
cat("Backbone density:", round(sum(mat_backbone > 0) / n_possible, 2), "\n")
```

```{r}
#| label: fig-hai-networks
#| fig-cap: "Full transition network vs disparity backbone"
#| fig-height: 6
#| fig-width: 14
par(mfrow = c(1, 2))
splot(mod, minimum = 0.01, layout = "fr", title = "Full Network (90%+ density)")
splot(mat_backbone, layout = "fr", title = "Backbone (p < 0.50)")
par(mfrow = c(1, 1))
```

## Robustness: Full vs Backbone

```{r}
#| label: fig-hai-robust
#| fig-cap: "Backbone extraction is essential for meaningful robustness analysis"
#| fig-height: 5
#| fig-width: 12
ggplot_robustness(
  "Full (no filtering)" = mat_full,
  "Backbone (p < 0.50)" = mat_backbone,
  n_iter = 100, seed = 42, ncol = 2
)
```

Without backbone extraction, all attack strategies overlap --- the dense network absorbs every removal. After filtering, clear differentiation emerges: betweenness attack collapses the network much faster than random removal.

## Comparing Human, AI, and Combined Networks

We apply the same disparity filter to separate Human and AI action networks:

```{r}
#| label: hai-compare
disp_human <- disparity_filter(tna(coding), level = 0.50)
disp_ai <- disparity_filter(tna(ai_coding), level = 0.50)

cat("Human backbone: ", disp_human$n_edges_filtered, "edges\n")
cat("AI backbone:    ", disp_ai$n_edges_filtered, "edges\n")
cat("Combined backbone:", disp_hai$n_edges_filtered, "edges\n")
```

```{r}
#| label: fig-hai-compare
#| fig-cap: "Robustness comparison across Human, AI, and Combined backbones"
#| fig-height: 5
#| fig-width: 16
ggplot_robustness(
  "Human" = disp_human$weights_filtered,
  "AI" = disp_ai$weights_filtered,
  "Combined" = mat_backbone,
  n_iter = 100, seed = 42, ncol = 3
)
```

## Robustness Metrics

```{r}
#| label: tbl-hai-summary
#| tbl-cap: "Human-AI Network: Full vs Backbone Robustness"
rob_hai_btw <- robustness(mat_backbone, measure = "betweenness")
rob_hai_deg <- robustness(mat_backbone, measure = "degree")
rob_hai_rnd <- robustness(mat_backbone, measure = "random", n_iter = 100, seed = 42)

rob_full_hai_btw <- robustness(mat_full, measure = "betweenness")
rob_full_hai_deg <- robustness(mat_full, measure = "degree")
rob_full_hai_rnd <- robustness(mat_full, measure = "random", n_iter = 100, seed = 42)

knitr::kable(rbind(
  cbind(Network = "Full", robustness_summary(rob_full_hai_btw, rob_full_hai_deg, rob_full_hai_rnd)),
  cbind(Network = "Backbone", robustness_summary(rob_hai_btw, rob_hai_deg, rob_hai_rnd))
), row.names = FALSE)
```


# Sequential vs Static Strategy

The `strategy` parameter controls whether centrality is recalculated after each removal or computed once on the original network.

```{r}
#| label: fig-strategy
#| fig-cap: "Sequential recalculation produces a stronger attack than static ordering"
#| fig-height: 5
#| fig-width: 12
# Sequential (default): recalculate after each removal
rob_seq_btw <- robustness(disp_student$weights_filtered,
                           measure = "betweenness", strategy = "sequential")
rob_seq_deg <- robustness(disp_student$weights_filtered,
                           measure = "degree", strategy = "sequential")
rob_seq_rnd <- robustness(disp_student$weights_filtered,
                           measure = "random", n_iter = 100, seed = 42)

# Static (brainGraph-style): compute centrality once
rob_sta_btw <- robustness(disp_student$weights_filtered,
                           measure = "betweenness", strategy = "static")
rob_sta_deg <- robustness(disp_student$weights_filtered,
                           measure = "degree", strategy = "static")

par(mfrow = c(1, 2))
plot_robustness(rob_seq_btw, rob_seq_deg, rob_seq_rnd,
                title = "Sequential (recalculate after each removal)")
plot_robustness(rob_sta_btw, rob_sta_deg, rob_seq_rnd,
                title = "Static (centrality computed once)")
par(mfrow = c(1, 1))
```

```{r}
#| label: tbl-strategy
#| tbl-cap: "AUC comparison: Sequential vs Static attack strategies"
knitr::kable(data.frame(
  Measure = c("Betweenness", "Degree"),
  Sequential = c(round(robustness_auc(rob_seq_btw), 3),
                 round(robustness_auc(rob_seq_deg), 3)),
  Static = c(round(robustness_auc(rob_sta_btw), 3),
             round(robustness_auc(rob_sta_deg), 3))
))
```

Sequential recalculation produces lower AUC (= more effective attack) because removing a hub changes which nodes are now most central. Static ordering uses the original ranking, which becomes stale as the network shrinks.


# Effect of Disparity Level

The disparity filter's `level` parameter controls how aggressively the backbone is pruned. Lower levels retain only the most statistically significant edges:

```{r}
#| label: fig-levels
#| fig-cap: "Stricter filtering reveals greater vulnerability"
levels_test <- c(0.01, 0.05, 0.10, 0.50)
colors_l <- c("#7570B3", "#D95F02", "#1B9E77", "#999999")

plot(NULL, xlim = c(0, 1), ylim = c(0, 1),
     xlab = "Fraction of removed nodes",
     ylab = "Fraction of remaining nodes",
     main = "Human-AI: Betweenness Attack at Different Backbone Levels",
     las = 1, xaxs = "i", yaxs = "i")
abline(h = seq(0, 1, 0.25), col = "gray90")
abline(v = seq(0, 1, 0.25), col = "gray90")

edge_counts <- vapply(levels_test, function(l) {
  disparity_filter(mod, level = l)$n_edges_filtered
}, integer(1))

for (i in seq_along(levels_test)) {
  disp <- disparity_filter(mod, level = levels_test[i])
  rob <- robustness(disp$weights_filtered, measure = "betweenness")
  lines(rob$removed_pct, rob$comp_pct, col = colors_l[i], lwd = 2)
}

legend("topright",
       legend = sprintf("p < %.2f (%d edges)", levels_test, edge_counts),
       col = colors_l, lwd = 2, bty = "n", title = "Disparity level")
```

Stricter levels produce sparser backbones that collapse faster under attack. We use `level = 0.50` throughout this tutorial, which provides a good balance between retaining meaningful structure and removing noise.


# Reproducibility

Random robustness is stochastic. Setting the `seed` parameter ensures identical results:

```{r}
#| label: reproducibility
rob_a <- robustness(g_weighted, measure = "random", n_iter = 100, seed = 123)
rob_b <- robustness(g_weighted, measure = "random", n_iter = 100, seed = 123)
identical(rob_a$comp_pct, rob_b$comp_pct)
```


# Input Types

`robustness()` and `disparity_filter()` accept all major network formats:

```{r}
#| label: input-types
# Matrix
rob_mat <- robustness(mat_backbone)

# igraph
rob_ig <- robustness(g_weighted)

# tna object
rob_tna <- robustness(mod)

# cograph_network
rob_cg <- robustness(as_cograph(mat_backbone))

cat("Matrix AUC: ", round(robustness_auc(rob_mat), 3), "\n")
cat("igraph AUC: ", round(robustness_auc(rob_ig), 3), "\n")
cat("tna AUC:    ", round(robustness_auc(rob_tna), 3), "\n")
cat("cograph AUC:", round(robustness_auc(rob_cg), 3), "\n")
```


# Function Reference

## Core Functions

| Function | Purpose |
|---|---|
| `disparity_filter(x, level)` | Extract statistically significant backbone |
| `robustness(x, type, measure, strategy)` | Simulate node/edge removal |
| `robustness_auc(x)` | Area under the robustness curve |
| `robustness_summary(...)` | AUC + critical thresholds |
| `plot_robustness(...)` | Base R robustness plot |
| `ggplot_robustness(...)` | ggplot2 faceted comparison |

## Key Parameters

| Parameter | Values | Default |
|---|---|---|
| `type` | `"vertex"`, `"edge"` | `"vertex"` |
| `measure` | `"betweenness"`, `"degree"`, `"random"` | `"betweenness"` |
| `strategy` | `"sequential"`, `"static"` | `"sequential"` |
| `n_iter` | Integer (random only) | `1000` |
| `seed` | Integer or `NULL` | `NULL` |
| `level` | Disparity filter significance | `0.05` |


# Key Takeaways

- **Backbone extraction matters**: Dense networks mask structural vulnerabilities. The disparity filter provides a statistically principled way to extract the significant skeleton.
- **Aggregation**: Minimal effect on vertex robustness, but substantial for edge robustness.
- **Strategy choice**: Sequential recalculation (default) produces stronger attacks than static ordering (brainGraph-style). Report which strategy you used.
- **Both preprocessing and strategy decisions should be reported** when publishing robustness results.


# References

Albert, R., Jeong, H., & Barabási, A.L. (2000). Error and attack tolerance of complex networks. *Nature*, 406, 378--382. doi:10.1038/35019019

Holme, P., Kim, B. J., Yoon, C. N., & Han, S. K. (2002). Attack vulnerability of complex networks. *Physical Review E*, 65(5), 056109. doi:10.1103/PhysRevE.65.056109

Serrano, M. Á., Boguñá, M., & Vespignani, A. (2009). Extracting the multiscale backbone of complex weighted networks. *Proceedings of the National Academy of Sciences*, 106(16), 6483--6488. doi:10.1073/pnas.0808904106
